Focasting The Trend of Vehicle Related Crime In Calgary In 2025\thanks{Code and data are available at: https://github.com/HaoboRrrr/calgary\_vehicle\_crime\_forcast}

### 2024-11-28 paper/paper.tex
small change to the paper

---
title: "Focasting The Trend of Vehicle Related Crime In Calgary In 2025"
subtitle: "An Analysis of Calgary's Crime Count From 2018 to 2024"
author: 
  - Haobo Ren
thanks: "Code and data are available at: https://github.com/HaoboRrrr/calgary_vehicle_crime_forcast"
date: 23-November-2024
date-format: long
abstract: "Vehicle-related crime is a long-standing challenge to urban security, so accurate prediction is necessary to effectively allocate police resources and develop security policies. This study examines trends in vehicle-related crime in Calgary, Canada, focusing on two types of crime: Theft of vehicle and Theft from vehicle. A Bayesian regression model is fit to historical crime data from January 2018 to October 2024, accounting for seasonal variations (months), long-term trends (years), and differences between crime categories. The usefulness of Bayesian modeling for crime trend forecasting is demonstrated by this analysis, providing law enforcement and policymakers practical insights to foresee and handle long-term and seasonal crime trends."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(here)
library(knitr)
library(arrow)
library(ggplot2)
library(bayesplot)

cleaned_data <- read_parquet(here("data/02-analysis_data/calgary_crime.parquet"))
fitted_data <- read_parquet(here("data/02-analysis_data/fitted_calgary_crime.parquet"))
```

# Introduction

@@

#Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

## Overview

The raw dataset obtained from Open Data Calgary[@OpenCalgary] was recorded and updated monthly by the Calgary Police Service. The data is considered cumulative as late-reported incidents are often received well after an offence has occurred. An incident is either reported just after the crime happened, or reported on the Calgary Police Service[@calgarypolice].

The data analysis and visualization is done in R[@citeR] with the following Packages: tidyverse[@tidy], janitor[@janitor], arrow[@arrow], rstanarm[@rstanarm], ggplot2[@ggplot2], dplyr[@dplyr], here, knitr.

## Measurement

@@

## Data Examination {#sec-dataexam}

@@

```{r}
#| echo: false
#| warning: false
raw_data <- read_csv(here("data/01-raw_data/raw_data.csv"))

# Get the column names of your raw data
column_names <- colnames(raw_data)

# Convert the column names to a data frame for kable
column_df <- as.data.frame(column_names)

# Display with kable
kable(column_df)
```
- Community: The dataset is spatially disaggregated into Calgary's various communities

@@

- Year: The year of each record indicate when the incidents were reported.

@@

- Category: Each entry is categorized into distinct crime types.

- month: The month of each incidents recorded. Range from 1-12 repeatedly each year

@@

- Crime Count: This variable quantifies the number of reported incidents of a specific crime type (e.g., theft from or of vehicles) within a given community and time period.

## Data Cleaning

Several steps was taken for better analysis of vehicle-related trend in Calgary. First, clean the names of all the columns. Second, the raw dataset contained granular records categorized by Community and specific details. These were aggregated to provide summarized counts of crimes (crime_count) by category, year, and month to align with analysis requirements. Third, a new time column was introduced in the cleaned data, combining year and month into a single date field (YYYY-MM-DD). This was essential for time-series analysis and visualization. Last, filter out the categories related to vehicle.

## Cleaned Data

The first 6 rows of the cleaned dataset are shown in @tbl-cleaned-data.

```{r}
#| label: tbl-cleaned-data
#| tbl-cap: Head of cleaned Calgary Crime data 
#| echo: false
#| warning: false

kable(head(cleaned_data))
```

## Response Variable

### Crime count of Theft FROM Vehicle

Summary statistics of Crime count of Theft FROM Vehicle are shown in the @tbl-response-theftfromvehicle

```{r}
#| label: tbl-response-theftfromvehicle
#| tbl-cap: Summary Statitic of crime_count(Theft FROM Vehicle)
#| echo: false
#| warning: false

kable(cleaned_data %>% filter(category == "Theft FROM Vehicle") %>% select(crime_count) %>% summary(crime_count))
```

The response variable, crime_count, represents the number of reported incidents of theft from vehicles within Calgary. The summary statistics reveal a significant variability in the crime counts, ranging from a minimum of 120 to a maximum of 1,619 reported cases. The median value is 916, indicating that half of the recorded months experienced fewer than 916 thefts, while the other half exceeded this value. The mean crime count, at 932.7, is slightly higher than the median, suggesting the presence of months with particularly high crime counts that may be skewing the average. The interquartile range (IQR), spanning from 788.2 (1st quartile) to 1,105.5 (3rd quartile), highlights the central tendency of crime occurrences, with most months falling within this range. These statistics emphasize the fluctuating nature of vehicle thefts in Calgary and provide a foundational understanding for modeling and forecasting future trends.

### Crime count of Theft OF Vehicle

Summary statistics of Crime count of Theft FROM Vehicle are shown in the @tbl-response-theftofvehicle

```{r}
#| label: tbl-response-theftofvehicle
#| tbl-cap: Summary Statitic of crime_count(Theft OF Vehicle)
#| echo: false
#| warning: false

kable(cleaned_data %>% filter(category == "Theft OF Vehicle") %>% select(crime_count) %>% summary(crime_count))
```

The response variable, crime_count, for Theft OF Vehicle incidents reflects the monthly reported counts of vehicles being stolen in Calgary. The dataset's summary statistics indicate that the crime counts range from a minimum of 146 to a maximum of 651 incidents. The median count is 424.5, suggesting that half of the months have fewer than 425 thefts, while the other half exceed this value. Interestingly, the mean crime count is 422.5, closely aligned with the median, suggesting a relatively symmetric distribution of the data. The interquartile range (IQR) spans from 361.2 (1st quartile) to 469.0 (3rd quartile), indicating that most monthly counts are concentrated within this range. These figures highlight the scale and regularity of vehicle theft in Calgary and form an essential basis for analyzing and forecasting trends in such crimes.

## Predictor Variable

### year

@@

### month and category

The variables month and category provide a detailed view of how vehicle-related crimes are distributed across time and crime types. Month captures seasonal variations, showing how crime rates fluctuate throughout the year. For example, there might be a tendency for higher crime counts during certain seasons, such as summer or winter, influenced by weather conditions or holiday activities. Category distinguishes between "Theft FROM Vehicle" and "Theft OF Vehicle," allowing us to compare the frequency and patterns of these two crime types. Together, these variables highlight differences in the nature and timing of vehicle-related crimes, offering valuable insights into their temporal and categorical distributions.

## Trend of The Two Crime

The trend of Theft FROM Vehicle and Theft OF Vehicle are shown in @fig-combined-trend

```{r}
#| label: fig-combined-trend
#| fig-cap: Trend of The Two Crime Combined
#| echo: false
#| warning: false
# Plot combined trends
ggplot(cleaned_data, aes(x = time, y = crime_count, color = category)) +
  geom_line() +
  labs(
    title = "Trend of Theft FROM Vehicle and Theft OF Vehicle",
    x = "Time",
    y = "Crime Count",
    color = "Category"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
\newpage

The trends for "Theft FROM Vehicle" and "Theft OF Vehicle" reveal distinct patterns over time. "Theft FROM Vehicle" consistently shows higher crime counts compared to "Theft OF Vehicle," indicating that it is a more prevalent issue. The trend for "Theft FROM Vehicle" demonstrates significant fluctuations, with peaks around 2019 and a general decline afterward, suggesting a reduction in these incidents in more recent years. In contrast, "Theft OF Vehicle" exhibits a relatively stable pattern with smaller variations over time. Although it remains less frequent than "Theft FROM Vehicle," its counts appear to decrease gradually, particularly after 2022. Both categories show declining trends toward the end of the observed period, which could reflect successful crime prevention measures or other external influences.

# Model

The Bayesian regression model was developed to analyze and forecast the trends of vehicle-related crimes in Calgary, focusing on the categories "Theft FROM Vehicle" and "Theft OF Vehicle." The primary goal of this modeling is to understand the relationship between crime counts and key predictors such as time (year and month) and crime categories, as well as their interactions. By incorporating temporal and categorical variables, the model seeks to capture long-term trends, seasonal patterns, and differences between the two types of crimes. Ultimately, the model provides a foundation for projecting future crime rates, offering valuable insights that can support strategic decision-making and the development of targeted crime prevention initiatives for 2025 and beyond.

Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

The Bayesian regression model is set up to examine the relationship between vehicle-related crime counts and the selected predictors. The response variable, $y_i$, represents the observed crime count for the i-th observation, where each observation corresponds to a specific month, year, and crime category. The predictors include a numeric variable for year, categorical variables for month and category, as well as interaction terms (category:month and category:year) to capture how crime patterns differ between the two crime types over time and across seasons.

@@

$$
y_i = \beta_0 + \beta_1 \text{year}_i + \beta_2 \text{month}_i + \beta_3 \text{category}_i + \beta_4 (\text{category} \times \text{month})_i + \beta_5 (\text{category} \times \text{year})_i + \epsilon_i
$$

Here, $\beta_0$ is the intercept, capturing the baseline level of crime counts. The coefficients $\beta_1$, $\beta_2$, and $\beta_3$ quantify the effects of year, month, and category, respectively, while $\beta_4$ and $\beta_5$ capture the interaction effects. The error term $\epsilon_i$ accounts for the residual variability in crime counts not explained by the predictors, modeled with a Gaussian distribution. This setup allows the model to estimate the relative contributions of each predictor to crime counts and their combined effects on the observed trends.

## Model justification

The primary goal of this model is to forecast trends in vehicle-related crimes in Calgary, focusing on two categories: "Theft FROM Vehicle" and "Theft OF Vehicle." To achieve this, a Bayesian regression model was chosen, as it provides a robust framework for incorporating prior knowledge, handling uncertainty, and capturing complex relationships between predictors.

The selection of predictors for the Bayesian regression model is grounded in their expected relationship with the response variable, crime_count, and their importance in explaining and forecasting trends in vehicle-related crimes. Below is a detailed justification for each predictor.

### year(Numeric Predictor)

@@

Relationship to Crime Counts: Crime counts for both categories ("Theft FROM Vehicle" and "Theft OF Vehicle") show clear temporal trends when plotted against year. For example, "Theft FROM Vehicle" demonstrates an overall decline in recent years, whereas "Theft OF Vehicle" shows a more gradual change (@fig-year-trend).

```{r}
#| label: fig-year-trend
#| tbl-cap: Trend of Crime Counts Over Years
#| echo: false
#| warning: false

ggplot(cleaned_data, aes(x = year, y = crime_count, color = category)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Trend of Crime Counts Over Years",
    x = "Year",
    y = "Crime Count",
    color = "Category"
  ) +
  theme_minimal()

```

### month (Categorical Predictor)

@@

### category (Categorical Predictor)

The category variable differentiates between "Theft FROM Vehicle" and "Theft OF Vehicle," two distinct types of crimes with different characteristics and frequencies. This variable is crucial for explaining the variation in crime counts, as the two categories exhibit different trends over time and across months.

Relationship to Crime Counts: "Theft FROM Vehicle" has consistently higher crime counts than "Theft OF Vehicle." Additionally, the trends for these categories vary significantly over time and seasons, as seen in prior visualizations. By including category as a predictor, the model can capture these differences and provide more accurate predictions.

### Interactions (Category:Month, Category:Year)

Including interaction terms between category and time variables (month and year) allows the model to capture how the seasonal and yearly trends differ for "Theft FROM Vehicle" and "Theft OF Vehicle." For example, one category might experience sharper seasonal peaks or steeper declines over the years compared to the other.

Relationship to Crime Counts: Visualizations of crime counts by year and month, separated by category, show evidence of these interactions. For instance, "Theft FROM Vehicle" exhibits more pronounced seasonal patterns and a sharper decline in recent years compared to "Theft OF Vehicle."

# Results

## Overview

@@

## Trends in Crime Counts

The actual and fitted crime trends for both categories from 2018 to 2025 are shown in @fig-forcast-trend. The fitted values closely match the actual values, indicating the model's strong predictive accuracy.

```{r}
#| label: fig-forcast-trend
#| fig-cap: Actual vs. Fitted Crime Trends (2018-2025)
#| echo: false
#| warning: false
ggplot(fitted_data, aes(x = time, y = crime_count, color = category, linetype = source)) +
  geom_line() +
  labs(
    title = "Actual vs. Fitted Crime Trends (2018-2025)",
    x = "Time",
    y = "Crime Count",
    color = "Category",
    linetype = "Source"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

## Regression Model Estimates

The regression model estimates, summarized in the @tbl-model-summary below, provide insight into how predictors such as year, month, and category influence vehicle-related crime counts in Calgary. The table includes coefficient estimates and their 95% confidence intervals to quantify the effect sizes and uncertainties associated with each predictor.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{float}
\usepackage{tabularray}
\usepackage[normalem]{ulem}
\usepackage{graphicx}
\UseTblrLibrary{booktabs}
\UseTblrLibrary{rotating}
\UseTblrLibrary{siunitx}
\NewTableCommand{\tinytableDefineColor}[3]{\definecolor{#1}{#2}{#3}}
\newcommand{\tinytableTabularrayUnderline}[1]{\underline{#1}}
\newcommand{\tinytableTabularrayStrikeout}[1]{\sout{#1}}
\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Focasting The Trend of Vehicle Related Crime In Calgary In 2025},
  pdfauthor={Haobo Ren},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Focasting The Trend of Vehicle Related Crime In Calgary In
2025\thanks{Code and data are available at:
https://github.com/HaoboRrrr/calgary\_vehicle\_crime\_forcast}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{An Analysis of Calgary's Crime Count From 2018 to 2024}
\author{Haobo Ren}
\date{November 23, 2024}

\begin{document}
\maketitle
\begin{abstract}
Vehicle-related crime is a long-standing challenge to urban security, so
accurate prediction is necessary to effectively allocate police
resources and develop security policies. This study examines trends in
vehicle-related crime in Calgary, Canada, focusing on two types of
crime: Theft of vehicle and Theft from vehicle. A Bayesian regression
model is fit to historical crime data from January 2018 to October 2024,
accounting for seasonal variations (months), long-term trends (years),
and differences between crime categories. The usefulness of Bayesian
modeling for crime trend forecasting is demonstrated by this analysis,
providing law enforcement and policymakers practical insights to foresee
and handle long-term and seasonal crime trends.
\end{abstract}

\section{Introduction}\label{introduction}

@@

\#Telegraphing paragraph: The remainder of this paper is structured as
follows. Section~\ref{sec-data}\ldots.

\section{Data}\label{sec-data}

\subsection{Overview}\label{overview}

The raw dataset obtained from Open Data Calgary(Calgary, n.d.) was
recorded and updated monthly by the Calgary Police Service. The data is
considered cumulative as late-reported incidents are often received well
after an offence has occurred. An incident is either reported just after
the crime happened, or reported on the Calgary Police Service(Service,
n.d.).

The data analysis and visualization is done in R(R Core Team 2023) with
the following Packages: tidyverse(Wickham et al. 2019), janitor(Firke
2023), arrow(Richardson et al. 2024), rstanarm(Goodrich et al. 2022),
ggplot2(Wickham 2016), dplyr(Wickham et al. 2022), here, knitr.

\subsection{Measurement}\label{measurement}

@@

\subsection{Data Examination}\label{sec-dataexam}

@@

\begin{longtable}[]{@{}l@{}}
\toprule\noalign{}
column\_names \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Community \\
Category \\
Crime Count \\
Year \\
Month \\
\end{longtable}

\begin{itemize}
\tightlist
\item
  Community: The dataset is spatially disaggregated into Calgary's
  various communities
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Year: The year of each record indicate when the incidents were
  reported.
\end{itemize}

@@

\begin{itemize}
\item
  Category: Each entry is categorized into distinct crime types.
\item
  month: The month of each incidents recorded. Range from 1-12
  repeatedly each year
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  Crime Count: This variable quantifies the number of reported incidents
  of a specific crime type (e.g., theft from or of vehicles) within a
  given community and time period.
\end{itemize}

\subsection{Data Cleaning}\label{data-cleaning}

Several steps was taken for better analysis of vehicle-related trend in
Calgary. First, clean the names of all the columns. Second, the raw
dataset contained granular records categorized by Community and specific
details. These were aggregated to provide summarized counts of crimes
(crime\_count) by category, year, and month to align with analysis
requirements. Third, a new time column was introduced in the cleaned
data, combining year and month into a single date field (YYYY-MM-DD).
This was essential for time-series analysis and visualization. Last,
filter out the categories related to vehicle.

\subsection{Cleaned Data}\label{cleaned-data}

The first 6 rows of the cleaned dataset are shown in
Table~\ref{tbl-cleaned-data}.

\begin{longtable}[]{@{}lrrrl@{}}

\caption{\label{tbl-cleaned-data}Head of cleaned Calgary Crime data}

\tabularnewline

\toprule\noalign{}
category & crime\_count & year & month & time \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Theft FROM Vehicle & 962 & 2018 & 1 & 2018-01-01 \\
Theft FROM Vehicle & 645 & 2018 & 2 & 2018-02-01 \\
Theft FROM Vehicle & 818 & 2018 & 3 & 2018-03-01 \\
Theft FROM Vehicle & 870 & 2018 & 4 & 2018-04-01 \\
Theft FROM Vehicle & 1063 & 2018 & 5 & 2018-05-01 \\
Theft FROM Vehicle & 1036 & 2018 & 6 & 2018-06-01 \\

\end{longtable}

\subsection{Response Variable}\label{response-variable}

\subsubsection{Crime count of Theft FROM
Vehicle}\label{crime-count-of-theft-from-vehicle}

Summary statistics of Crime count of Theft FROM Vehicle are shown in the
Table~\ref{tbl-response-theftfromvehicle}

\begin{longtable}[]{@{}ll@{}}

\caption{\label{tbl-response-theftfromvehicle}Summary Statitic of
crime\_count(Theft FROM Vehicle)}

\tabularnewline

\toprule\noalign{}
& crime\_count \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
& Min. : 120.0 \\
& 1st Qu.: 788.2 \\
& Median : 916.0 \\
& Mean : 932.7 \\
& 3rd Qu.:1105.5 \\
& Max. :1619.0 \\

\end{longtable}

The response variable, crime\_count, represents the number of reported
incidents of theft from vehicles within Calgary. The summary statistics
reveal a significant variability in the crime counts, ranging from a
minimum of 120 to a maximum of 1,619 reported cases. The median value is
916, indicating that half of the recorded months experienced fewer than
916 thefts, while the other half exceeded this value. The mean crime
count, at 932.7, is slightly higher than the median, suggesting the
presence of months with particularly high crime counts that may be
skewing the average. The interquartile range (IQR), spanning from 788.2
(1st quartile) to 1,105.5 (3rd quartile), highlights the central
tendency of crime occurrences, with most months falling within this
range. These statistics emphasize the fluctuating nature of vehicle
thefts in Calgary and provide a foundational understanding for modeling
and forecasting future trends.

\subsubsection{Crime count of Theft OF
Vehicle}\label{crime-count-of-theft-of-vehicle}

Summary statistics of Crime count of Theft FROM Vehicle are shown in the
Table~\ref{tbl-response-theftofvehicle}

\begin{longtable}[]{@{}ll@{}}

\caption{\label{tbl-response-theftofvehicle}Summary Statitic of
crime\_count(Theft OF Vehicle)}

\tabularnewline

\toprule\noalign{}
& crime\_count \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
& Min. :146.0 \\
& 1st Qu.:361.2 \\
& Median :424.5 \\
& Mean :422.5 \\
& 3rd Qu.:469.0 \\
& Max. :651.0 \\

\end{longtable}

The response variable, crime\_count, for Theft OF Vehicle incidents
reflects the monthly reported counts of vehicles being stolen in
Calgary. The dataset's summary statistics indicate that the crime counts
range from a minimum of 146 to a maximum of 651 incidents. The median
count is 424.5, suggesting that half of the months have fewer than 425
thefts, while the other half exceed this value. Interestingly, the mean
crime count is 422.5, closely aligned with the median, suggesting a
relatively symmetric distribution of the data. The interquartile range
(IQR) spans from 361.2 (1st quartile) to 469.0 (3rd quartile),
indicating that most monthly counts are concentrated within this range.
These figures highlight the scale and regularity of vehicle theft in
Calgary and form an essential basis for analyzing and forecasting trends
in such crimes.

\subsection{Predictor Variable}\label{predictor-variable}

\subsubsection{year}\label{year}

@@

\subsubsection{month and category}\label{month-and-category}

The variables month and category provide a detailed view of how
vehicle-related crimes are distributed across time and crime types.
Month captures seasonal variations, showing how crime rates fluctuate
throughout the year. For example, there might be a tendency for higher
crime counts during certain seasons, such as summer or winter,
influenced by weather conditions or holiday activities. Category
distinguishes between ``Theft FROM Vehicle'' and ``Theft OF Vehicle,''
allowing us to compare the frequency and patterns of these two crime
types. Together, these variables highlight differences in the nature and
timing of vehicle-related crimes, offering valuable insights into their
temporal and categorical distributions.

\subsection{Trend of The Two Crime}\label{trend-of-the-two-crime}

The trend of Theft FROM Vehicle and Theft OF Vehicle are shown in
Figure~\ref{fig-combined-trend}

\begin{figure}

\centering{

\includegraphics{paper_files/figure-pdf/fig-combined-trend-1.pdf}

}

\caption{\label{fig-combined-trend}Trend of The Two Crime Combined}

\end{figure}%

The trends for ``Theft FROM Vehicle'' and ``Theft OF Vehicle'' reveal
distinct patterns over time. ``Theft FROM Vehicle'' consistently shows
higher crime counts compared to ``Theft OF Vehicle,'' indicating that it
is a more prevalent issue. The trend for ``Theft FROM Vehicle''
demonstrates significant fluctuations, with peaks around 2019 and a
general decline afterward, suggesting a reduction in these incidents in
more recent years. In contrast, ``Theft OF Vehicle'' exhibits a
relatively stable pattern with smaller variations over time. Although it
remains less frequent than ``Theft FROM Vehicle,'' its counts appear to
decrease gradually, particularly after 2022. Both categories show
declining trends toward the end of the observed period, which could
reflect successful crime prevention measures or other external
influences.

\section{Model}\label{model}

The Bayesian regression model was developed to analyze and forecast the
trends of vehicle-related crimes in Calgary, focusing on the categories
``Theft FROM Vehicle'' and ``Theft OF Vehicle.'' The primary goal of
this modeling is to understand the relationship between crime counts and
key predictors such as time (year and month) and crime categories, as
well as their interactions. By incorporating temporal and categorical
variables, the model seeks to capture long-term trends, seasonal
patterns, and differences between the two types of crimes. Ultimately,
the model provides a foundation for projecting future crime rates,
offering valuable insights that can support strategic decision-making
and the development of targeted crime prevention initiatives for 2025
and beyond.

Background details and diagnostics are included in
\textbf{?@sec-model-details}.

\subsection{Model set-up}\label{model-set-up}

The Bayesian regression model is set up to examine the relationship
between vehicle-related crime counts and the selected predictors. The
response variable, \(y_i\), represents the observed crime count for the
i-th observation, where each observation corresponds to a specific
month, year, and crime category. The predictors include a numeric
variable for year, categorical variables for month and category, as well
as interaction terms (category:month and category:year) to capture how
crime patterns differ between the two crime types over time and across
seasons.

@@

\[
y_i = \beta_0 + \beta_1 \text{year}_i + \beta_2 \text{month}_i + \beta_3 \text{category}_i + \beta_4 (\text{category} \times \text{month})_i + \beta_5 (\text{category} \times \text{year})_i + \epsilon_i
\]

Here, \(\beta_0\) is the intercept, capturing the baseline level of
crime counts. The coefficients \(\beta_1\), \(\beta_2\), and \(\beta_3\)
quantify the effects of year, month, and category, respectively, while
\(\beta_4\) and \(\beta_5\) capture the interaction effects. The error
term \(\epsilon_i\) accounts for the residual variability in crime
counts not explained by the predictors, modeled with a Gaussian
distribution. This setup allows the model to estimate the relative
contributions of each predictor to crime counts and their combined
effects on the observed trends.

\subsection{Model justification}\label{model-justification}

The primary goal of this model is to forecast trends in vehicle-related
crimes in Calgary, focusing on two categories: ``Theft FROM Vehicle''
and ``Theft OF Vehicle.'' To achieve this, a Bayesian regression model
was chosen, as it provides a robust framework for incorporating prior
knowledge, handling uncertainty, and capturing complex relationships
between predictors.

The selection of predictors for the Bayesian regression model is
grounded in their expected relationship with the response variable,
crime\_count, and their importance in explaining and forecasting trends
in vehicle-related crimes. Below is a detailed justification for each
predictor.

\subsubsection{year(Numeric Predictor)}\label{yearnumeric-predictor}

@@

Relationship to Crime Counts: Crime counts for both categories (``Theft
FROM Vehicle'' and ``Theft OF Vehicle'') show clear temporal trends when
plotted against year. For example, ``Theft FROM Vehicle'' demonstrates
an overall decline in recent years, whereas ``Theft OF Vehicle'' shows a
more gradual change (Figure~\ref{fig-year-trend}).

\begin{figure}

\centering{

\includegraphics{paper_files/figure-pdf/fig-year-trend-1.pdf}

}

\caption{\label{fig-year-trend}}

\end{figure}%

\subsubsection{month (Categorical
Predictor)}\label{month-categorical-predictor}

@@

\subsubsection{category (Categorical
Predictor)}\label{category-categorical-predictor}

The category variable differentiates between ``Theft FROM Vehicle'' and
``Theft OF Vehicle,'' two distinct types of crimes with different
characteristics and frequencies. This variable is crucial for explaining
the variation in crime counts, as the two categories exhibit different
trends over time and across months.

Relationship to Crime Counts: ``Theft FROM Vehicle'' has consistently
higher crime counts than ``Theft OF Vehicle.'' Additionally, the trends
for these categories vary significantly over time and seasons, as seen
in prior visualizations. By including category as a predictor, the model
can capture these differences and provide more accurate predictions.

\subsubsection{Interactions (Category:Month,
Category:Year)}\label{interactions-categorymonth-categoryyear}

Including interaction terms between category and time variables (month
and year) allows the model to capture how the seasonal and yearly trends
differ for ``Theft FROM Vehicle'' and ``Theft OF Vehicle.'' For example,
one category might experience sharper seasonal peaks or steeper declines
over the years compared to the other.

Relationship to Crime Counts: Visualizations of crime counts by year and
month, separated by category, show evidence of these interactions. For
instance, ``Theft FROM Vehicle'' exhibits more pronounced seasonal
patterns and a sharper decline in recent years compared to ``Theft OF
Vehicle.''

\section{Results}\label{results}

\subsection{Overview}\label{overview-1}

@@

\subsection{Trends in Crime Counts}\label{trends-in-crime-counts}

The actual and fitted crime trends for both categories from 2018 to 2025
are shown in Figure~\ref{fig-forcast-trend}. The fitted values closely
match the actual values, indicating the model's strong predictive
accuracy.

\begin{figure}

\centering{

\includegraphics{paper_files/figure-pdf/fig-forcast-trend-1.pdf}

}

\caption{\label{fig-forcast-trend}Actual vs.~Fitted Crime Trends
(2018-2025)}

\end{figure}%

@@

\subsection{Regression Model
Estimates}\label{regression-model-estimates}

The regression model estimates, summarized in the
Table~\ref{tbl-model-summary} below, provide insight into how predictors
such as year, month, and category influence vehicle-related crime counts
in Calgary. The table includes coefficient estimates and their 95\%
confidence intervals to quantify the effect sizes and uncertainties
associated with each predictor.

\begin{table}

\caption{\label{tbl-model-summary}Regression Estimates for Crime Counts}

\centering{
